---
title: "Irec and creel data exploration"
author: "Catarina Wor"
date:  "`r format(Sys.time(), '%B %Y')`"
output: github_document
---



# Steps
 - Reproduce current simple model
 - compare with updated version including area



```{r load, include=FALSE}

library(ggplot2)
library(dplyr)
library(grid)
library(glmmTMB)
library(DHARMa)
library(brms)


#rmarkdown::render("irec_summary.Rmd")

#read in data
source(here::here("R","format-data.R"))
dat <- format_data()
dat$creel_orig <- dat$creel
dat$creel<- round(dat$creel)
dat <- format_data()

```

The original model is given by:

$$
Creel_i =  \beta_1\cdot z_i + \epsilon_i
$$
$$
z_i = iREC_i   + \nu_{i} 
$$

$$
\epsilon_i \sim N(0,\sigma_{\epsilon})
$$
$$
\nu_{i} \sim N(0,\sigma_{\nu})
$$


This model is run individually for each year, and disposition (kept/released). In the initial runs of this project, I thought of grouping all the years as replicates, however that would mean that the catch estimates for year Y would change every time the calibration is re-run, which is not ideal and may cause a lot of confusion. 

So I am reproducing the analysis (kind of), by year. Using year 2019 as an example here.



##  Reproduce default model

This is an attempt to reproduce the original model using glmmTMB - So I can compare it with the model with area as a covariate. The original code is written in JAGS, and takes into account the known sd of the observed irec estimates. -- I am not sure of how to reproduce that in glmmTMB, so I crated a dummy variable for each observation.


```{r filterdat, include=FALSE}
#let's exclude 0 estimates of iRec

dat_pos <- filter(dat, irec > 0 & year==2019)
dat_pos$dum<-seq_len(nrow(dat_pos))

```

```{r model0, include=TRUE}

m <- glmmTMB(creel ~ 0+ (0+irec|dum),
             data = dat_pos)
summary(m)

#diagnostics
bm <- DHARMa::simulateResiduals(m, n = 1000)
plotResiduals(bm, form = dat_pos$irec)

```

## Updated method

This update is based on Sean's previous work on the explore.R file. To my suprise, the best model still performs well when the data is split by year. 

```{r model2, include=TRUE}

dat_pos$log_irec <- log(dat_pos$irec) 

m1.2 <- glmmTMB(creel ~ 1 + log_irec +  (1 + log_irec | region), 
                data = dat_pos, family = nbinom1(), dispformula = ~log_irec)

summary(m1.2)

#diagnostics
r <- DHARMa::simulateResiduals(m1.2, n = 1000)
plot(r)

```

## brms

I am trying to reproduce the model from glmmTMB in brms but unsure if this one makes sense. 


```{r brms, include=TRUE}

#compare models with and without the cv 
dat_pos$log_irec_cent <- dat_pos$log_irec - mean(dat_pos$log_irec)
dat_pos$irec_cv <- dat_pos$sdirec/ dat_pos$irec

fit1 <- brm(
  bf(
    creel ~   me(log_irec_cent,irec_cv) +
      (me(log_irec_cent,irec_cv) | region), 
    shape ~ log_irec_cent), 
  data = dat_pos, 
  family = negbinomial(),
  iter = 600, chains = 2, cores = 2
)

fit1

nd1 <- expand.grid(   
  region = unique(dat_pos$region),
  log_irec_cent = seq(min(dat_pos$log_irec_cent), max(dat_pos$log_irec_cent), length.out = 100),
  irec_cv = mean(dat_pos$irec_cv))
  #seq(min(dat_pos$irec_cv), max(dat_pos$irec_cv), length.out = 100))
 

nd1$log_irec <- nd1$log_irec_cent + mean(dat_pos$log_irec)
nd1$irec <- exp(nd1$log_irec)

x1 <- tidybayes::add_linpred_draws(nd1, fit1, ndraws = 100, transform = TRUE)
  
  ggplot(data=x1) +
  geom_line(aes(irec, .linpred, group = paste(region, .draw), colour = region),alpha = 0.2) +
  geom_point(data=dat_pos,aes(x=irec,y=creel))+
  facet_wrap(~region) +
  scale_x_log10() +
  scale_y_log10()

```



