---
title: "Irec and creel initial analysis and data exploration"
author: "Catarina Wor"
date:  "`r format(Sys.time(), '%B %Y')`"
output: bookdown::html_document2
---

 

```{r load, include=FALSE}

library(ggplot2)
library(dplyr)
library(grid)
library(brms)
#rmarkdown::render("irec_creel.Rmd")

#read in data
creel<-read.csv("../data/creel_filter.csv")
irec<-read.csv("../data/iRecchinook_2012_2021.csv")

arealu <- read.csv("../data/areaLU.csv") 


```



# Data groupings decisions

Based on previus explorations of dtat distributions  the following decisions were made:

- Exclude year 2012, because it is the first year in which iRec became operational and the data distribution for that year looked visibly different from that for orther years. 
- Group areas by logistical area following the lu_grouping3 provided by Kris Hein, which group all the PFMAs into `sum(!is.na(unique(arealu$LU_GROUPING3)))` areas. 
- Inspection of the creel quality control did not seem to show a lot of difference in the distribution of the data. For now we will opt to include all creel data.
- iRec data is inputed: any missing observations are inputed with 0's. If no survey respondant said they caught a chinook, then our estimate is 0. 
- Months were aggregates into offseason (October-April) and peakseanon(May-September).


```{r wrangleirec, include=FALSE}

creel<- creel%>%
      rename(AREA=PFMA)%>%left_join(arealu[,c("AREA","LU_GROUPING3")])%>%
      filter(YEAR>2012)

#input 0s for missing observations on irec
#create df for all possible observations - using variables of interest only
#if other variables are considered, need to include them here
allobs <- expand.grid(list(AREA=unique(irec$AREA),
  YEAR=unique(irec$YEAR),
  MONTH=unique(irec$MONTH),
  DISPOSITION=unique(irec$DISPOSITION)))

#create zero observations and remove in river fisheries
irecall<-left_join(allobs,irec) %>%
filter(AREA != "Area 29 (In River)", YEAR>2012)

irecall$ESTIMATE[is.na(irecall$ESTIMATE)]<-0

irecall<-irecall%>%left_join(arealu[,c("AREA","LU_GROUPING3")])


irecc<- irecall %>%
  #rename(IREC=ESTIMATE)%>%
  select(c(AREA,YEAR,MONTH,DISPOSITION,ESTIMATE,VARIANCE,LU_GROUPING3))%>%
  group_by( AREA,YEAR,MONTH,DISPOSITION,LU_GROUPING3) %>% summarise(ESTIMATE = sum(ESTIMATE), VARIANCE = sum(VARIANCE))%>%
  mutate(SD=sqrt(VARIANCE))%>%
  select(c(!VARIANCE))%>%
  mutate(SURVEY="iRec")


```

```{r wranglecreel, include=FALSE}


creelf<- creel %>%
  filter(ESTIMATE_SOURCE=="Creel")%>%
  mutate(SURVEY=case_when(
  Include..20.=="Y"~ "creel20",
  Include..15.=="Y"~ "creel15",
  TRUE ~ "creel"))%>%
  rename(SD=STANDARD_ERROR,DISPOSITION=TYPE)%>%
  select(c(AREA,YEAR,MONTH,DISPOSITION,ESTIMATE,SD, SURVEY, LU_GROUPING3))



creelcc<-creelf%>%
rename(CREEL=ESTIMATE,SDCREEL=SD)


ireccc<-irecc%>%
rename(IREC=ESTIMATE,SDIREC=SD)%>%
  select(c(!SURVEY))


datxy<-left_join(ireccc,creelcc) %>%
mutate(SEASON=if_else(MONTH<5|MONTH>9,"offseason","peakseason"))



```




# Relationships between creel and irec

## Relationship by month


```{r relbymonth, echo=FALSE, fig.height = 8, fig.width = 14, fig.cap = "Relationship between iRec and Creel estimates by  month and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION), size=2,alpha=.5)
p <- p + facet_wrap(~SEASON, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")
p <- p + geom_abline(slope=1,intercept=0)
p

p <- ggplot(datxy,aes(x=CREEL, y=IREC,color=as.factor(MONTH)))
p <- p + geom_point(size=2, alpha=.5)
p <- p + geom_smooth(method = lm, se     = FALSE, size   = 1, alpha  = .8) # to add regression line
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.9,option = "D")
p <- p + theme(legend.position="bottom")
p


```

# Relationship by area

```{r byarea, echo=FALSE, fig.height = 13, fig.width = 13, fig.cap = "Relationship between iRec and Creel estimaes by  area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION),size=2, alpha=.5)
p <- p + facet_wrap(~LU_GROUPING3, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")
#p <- p + coord_cartesian(ylim=c(0,max(datxy$CREEL,datxy$IREC, na.rm=T)),xlim=c(0,max(datxy$CREEL,datxy$IREC,na.rm=T)))
p <- p + geom_abline(slope=1,intercept=0)
p <- p + theme(legend.position="bottom")
p



p <- ggplot(datxy,aes(x=CREEL, y=IREC,color=LU_GROUPING3))
p <- p + geom_point(size=2, alpha=.5)
p <- p + geom_smooth(method = lm, se     = FALSE, size   = 1, alpha  = .8) # to add regression line
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")
p <- p + theme(legend.position="bottom")
p
```

# Relationship by year

```{r byyear, echo=FALSE, fig.height = 13, fig.width = 13, fig.cap = "Relationship between iRec and Creel estimaes by  area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION),size=2, alpha=.5)
p <- p + facet_wrap(~YEAR, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")

p <- p + geom_abline(slope=1,intercept=0)
p <- p + theme(legend.position="bottom")
p
```


## Look at proportions of zeroes in the data

Proportion of zeroes in irec is `sum(datxy$IREC==0)/nrow(datxy)`

```{r propzero,include=FALSE}

prop0 <-function(x){sum(x==0)/length(x)}

aggregate(datxy$IREC,by=list(datxy$MONTH),prop0)
aggregate(datxy$IREC,by=list(datxy$LU_GROUPING3),prop0)
aggregate(datxy$IREC,by=list(datxy$YEAR),prop0)

sum(datxy$IREC==0)/nrow(datxy)





```

```{r echo=FALSE,results = 'asis'}

prop0 <-function(x){sum(x==0)/length(x)}

knitr::kable(aggregate(datxy$IREC,by=list(datxy$MONTH),prop0), caption="proportion of zeroes in irec by month ")

```

```{r echo=FALSE,results = 'asis'}

prop0 <-function(x){sum(x==0)/length(x)}

knitr::kable(aggregate(datxy$IREC,by=list(datxy$LU_GROUPING3),prop0), caption="proportion of zeroes in irec by area ")

```

```{r echo=FALSE,results = 'asis'}

prop0 <-function(x){sum(x==0)/length(x)}

knitr::kable(aggregate(datxy$IREC,by=list(datxy$YEAR),prop0), caption="proportion of zeroes in irec by year ")

```


## Initial analysis

Questions:

- How to inform the brm function of the SD estimates for both IREC and Creel variables? the se option for the response variable is only available for normal, student t and skewed normal options ( I changed the code below just for testing. I am wondering if I could use weights() and weight observations by 1/SDIREC, not the same as including a latent variable though. 
- How do I suppress the intercept? We want the model to have a 0 intercept. 
- How would I consider area (LU_GROUPING3) a hierarchical variable?  
- There are quite a few missing values in the explanatory variables. I think these points are being dropped from the model fit. But I may want to predict for those values later, how would I do that?
- How to add the SD estimates for creel?
- Priors -- I still need to look into that, any suggestions are welcome.


```{r test model,include=FALSE}


# This model is just a first attempt to get brm working
# Do not run when rendering as the results are not meaningful and it takes a loooong time to run
 fit1 <- brm(formula= IREC |se(SDIREC,sigma = TRUE) ~ -1 + CREEL + SEASON + (-1 + CREEL |LU_GROUPING3), 
  data=datxy, family=gaussian)


#stancode(fit1)
#sdata <- standata(fit1)
#names(sdata)

### obtain model summaries and plots
#summary(fit1, waic = TRUE)
#plot(fit1, ask = FALSE)
#plot(marginal_effects(fit1), ask = FALSE)


```





# Background

We propose to develop a calibration model to combine both survey methods to a single recreational catch estimate, taking into account month and area. 

I am thinking this would be a state-space mode as we have error associated with both surveys, something like:


The simple model (one currently being used): 

\begin{align}
&y_i =  \beta_1\cdot z_i + \epsilon_i\\
&z_i = Creel_i   + \nu_{i} \\
&y_i = iREC_i   + \omega_{i} 
\end{align}

\begin{align}
&\epsilon_i \sim N(0,\sigma_{\epsilon})\\
&\nu_{i} \sim N(0,\sigma_{\nu})\\
&\omega_{i} \sim N(0,\sigma_{\omega i})
\end{align}

 where:

$iREC_i$ =  iREC observation 

$Creel_i$ =  creel observation 

$\beta_1$ =  slope associated with $Creel_i$ 

$z_i$ = smoothed ('true') $Creel_i$ 

$y_i$ = smoothed ('true') $iREC_i$ 

$\epsilon_i$ and $\nu_{i}$ are random normal variables 


Adding area and month:

\begin{align}
&y_i =  \beta_1\cdot z_i + \beta_2\cdot area_i + \beta_3\cdot season_i + \epsilon_i\\
&z_i = Creel_i + \nu_{i} \\
&y_i = iREC_i   + \omega_{i} 
\end{align}

\begin{align}
&\epsilon_i \sim Distribution(mean,variance)\\
&\nu_{i} \sim N(0,\sigma_{\nu i})\\
&\omega_{i} \sim N(0,\sigma_{\omega i})
\end{align}

 where: 
$\beta_2$ = slope associated with area 



$area_i$ = area for each observation (categorical) 

$\beta_3$ = slope associated with month 

$month_i$ = month for each observation (cyclic variable?) 

## Questions about model formulation:


