---
title: "Irec and creel initial analysis and data exploration"
author: "Catarina Wor"
date:  "`r format(Sys.time(), '%B %Y')`"
output: bookdown::html_document2
---

 

```{r load, include=FALSE}

library(ggplot2)
library(dplyr)
library(grid)
#rmarkdown::render("irec_summary.Rmd")

#read in data
creel<-read.csv("../data/creel_filter.csv")
irec<-read.csv("../data/iRecchinook_2012_2021.csv")

arealu <- read.csv("../data/areaLU.csv") 


```



# Data groupings decisions

Based on previus explorations of dtat distributions  the following decisions were made:

- Exclude year 2012, because it is the first year in which iRec became operational and the data distribution for that year looked visibly different from that for orther years. 
- Group areas by logistical area following the lu_grouping3 provided by Kris Hein, which group all the PFMAs into `sum(!is.na(unique(arealu$LU_GROUPING3)))` areas. 
- Inspection of the creel quality control did not seem to show a lot of difference in the distribution of the data. For now we will opt to include all creel data.
- iRec data is inputed: any missing observations are inputed with 0's. If no survey respondant said they caught a chinook, then our estimate is 0. 
- Months were aggregates into offseason (October-April) and peakseanon(May-September).


```{r wrangleirec, include=FALSE}

creel<- creel%>%
      rename(AREA=PFMA)%>%left_join(arealu[,c("AREA","LU_GROUPING3")])%>%
      filter(YEAR>2012)

#input 0s for missing observations on irec
#create df for all possible observations - using variables of interest only
#if other variables are considered, need to include them here
allobs <- expand.grid(list(AREA=unique(irec$AREA),
  YEAR=unique(irec$YEAR),
  MONTH=unique(irec$MONTH),
  DISPOSITION=unique(irec$DISPOSITION)))

#create zero observations and remove in river fisheries
irecall<-left_join(allobs,irec) %>%
filter(AREA != "Area 29 (In River)", YEAR>2012)

irecall$ESTIMATE[is.na(irecall$ESTIMATE)]<-0

irecall<-irecall%>%left_join(arealu[,c("AREA","LU_GROUPING3")])


irecc<- irecall %>%
  #rename(IREC=ESTIMATE)%>%
  select(c(AREA,YEAR,MONTH,DISPOSITION,ESTIMATE,VARIANCE,LU_GROUPING3))%>%
  group_by( AREA,YEAR,MONTH,DISPOSITION,LU_GROUPING3) %>% summarise(ESTIMATE = sum(ESTIMATE), VARIANCE = sum(VARIANCE))%>%
  mutate(SD=sqrt(VARIANCE))%>%
  select(c(!VARIANCE))%>%
  mutate(SURVEY="iRec")


```

```{r wranglecreel, include=FALSE}


creelf<- creel %>%
  filter(ESTIMATE_SOURCE=="Creel")%>%
  mutate(SURVEY=case_when(
  Include..20.=="Y"~ "creel20",
  Include..15.=="Y"~ "creel15",
  TRUE ~ "creel"))%>%
  rename(SD=STANDARD_ERROR,DISPOSITION=TYPE)%>%
  select(c(AREA,YEAR,MONTH,DISPOSITION,ESTIMATE,SD, SURVEY, LU_GROUPING3))



creelcc<-creelf%>%
rename(CREEL=ESTIMATE,SDCREEL=SD)


ireccc<-irecc%>%
rename(IREC=ESTIMATE,SDIREC=SD)%>%
  select(c(!SURVEY))


datxy<-left_join(ireccc,creelcc) %>%
mutate(SEASON=if_else(MONTH<5|MONTH>9,"offseason","peakseason"))




```




# Relationships between creel and irec

## Relationship by month


```{r relbymonth, echo=FALSE, fig.height = 8, fig.width = 14, fig.cap = "Relationship between iRec and Creel estimates by  month and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION), size=2,alpha=.5)
p <- p + facet_wrap(~SEASON, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")
p <- p + geom_abline(slope=1,intercept=0)
p


```

# Relationship by area

```{r byarea, echo=FALSE, fig.height = 13, fig.width = 13, fig.cap = "Relationship between iRec and Creel estimaes by  area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION),size=2, alpha=.5)
p <- p + facet_wrap(~LU_GROUPING3, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")
#p <- p + coord_cartesian(ylim=c(0,max(datxy$CREEL,datxy$IREC, na.rm=T)),xlim=c(0,max(datxy$CREEL,datxy$IREC,na.rm=T)))
p <- p + geom_abline(slope=1,intercept=0)
p <- p + theme(legend.position="bottom")
p
```

# Relationship by year

```{r byyear, echo=FALSE, fig.height = 13, fig.width = 13, fig.cap = "Relationship between iRec and Creel estimaes by  area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

p <- ggplot(datxy)
p <- p + geom_point(aes(x=CREEL, y=IREC,color=DISPOSITION, shape=DISPOSITION),size=2, alpha=.5)
p <- p + facet_wrap(~YEAR, scales="free")
p <- p + theme_bw(16)+labs( x="CREEL", y="iREC")
p <- p + scale_color_viridis_d(end = 0.8,option = "C")

p <- p + geom_abline(slope=1,intercept=0)
p <- p + theme(legend.position="bottom")
p
```


## Look at proportions of zeroes in the data


```{r propzero,include=FALSE}

prop0 <-function(x){sum(x==0)/length(x)}

aggregate(datxy$IREC,by=list(datxy$MONTH),prop0)
aggregate(datxy$IREC,by=list(datxy$LU_GROUPING3),prop0)
aggregate(datxy$IREC,by=list(datxy$YEAR),prop0)

sum(datxy$IREC==0)/nrow(datxy)





```





## Initial analysis

Questions:

- How to inform the brm function of the SD estimates for both IREC and Creel variables? the se option for the response variable is only available for normal, student t and skewed normal options ( I changed the code below just for testing. I am wondering if I could use weights() and weight observations by 1/SDIREC, not the same as including a latent variable though. 
- How do I suppress the intercept? We want the model to have a 0 intercept. 
- There are quite a few missing values in the explanatory variables. I think these points are being dropped from the model fit. But I may want to predict for those values later, how would I do that?
- How would I consider area (LU_GROUPING3) a hierarchical variable?  


```{r propzero,include=FALSE}


# This model is just a firs attempt
fit1 <- brm(formula= IREC |se(SDIREC,sigma = TRUE) ~ CREEL|LU_GROUPING3 + SEASON,
  data=datxy, family=gaussian)


stancode(fit1)
sdata <- standata(fit1)
names(sdata)

### obtain model summaries and plots
summary(fit1, waic = TRUE)
plot(fit1, ask = FALSE)
plot(marginal_effects(fit1), ask = FALSE)


```






## iRec:creel ratio by month


```{r ratbymonth, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = " Ratio between iRec and Creel estimates by  month and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}

datxy$ratio <- datxy$IREC/datxy$CREEL
datxy <- datxy[is.finite(datxy$ratio),] 


 

p <- ggplot(datxy, aes(x=as.factor(MONTH),y=ratio, color=DISPOSITION))
p <- p + geom_boxplot(width=1, position = position_dodge(width = 1))
p <- p +geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="month", y="iRec:creel")




p2 <- ggplot(datxy, aes(x=as.factor(MONTH),y=ratio, color=DISPOSITION))
p2 <- p2 + geom_boxplot(width=1, position = position_dodge(width = 1))
p2 <- p2 +geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p2 <- p2 + scale_color_viridis_d(end = 0.8,option = "A")
p2 <- p2 + theme_bw(16)+labs( x="month", y="iRec:creel")
p2 <- p2 + coord_cartesian( ylim = c(0,10))



grid.newpage()
grid.draw(rbind(ggplotGrob(p), ggplotGrob(p2), size = "last"))

```


## iRec:creel ratio by area

```{r ratbyarea, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = "Ratio between iRec and Creel estimates by area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}


p <- ggplot(datxy, aes(x=as.factor(LU_GROUPING3),y=ratio, color=DISPOSITION))
p <- p + geom_boxplot(width=1, position = position_dodge(width = 1))
p <- p +geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="Area", y="iRec:creel")
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))


p2 <- ggplot(datxy, aes(x=as.factor(LU_GROUPING3),y=ratio, color=DISPOSITION))
p2 <- p2 + geom_boxplot(width=1, position = position_dodge(width = 1))
p2 <- p2 +geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p2 <- p2 + scale_color_viridis_d(end = 0.8,option = "A")
p2 <- p2 + theme_bw(16)+labs( x="Area", y="iRec:creel")
p2 <- p2 + theme(axis.text.x = element_text(angle = 45, hjust=1))
p2 <- p2 + coord_cartesian( ylim = c(0,15))



grid.newpage()
grid.draw(rbind(ggplotGrob(p), ggplotGrob(p2), size = "last"))

```


**The year figure seems to suggest that 2012 and 2013 were different from all the other years, Perhaps people were still getting used to irec?**

```{r ratbyyear, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = "Ratio between iRec and Creel estimates by  area and disposition (kept or released) for the year of 2020. Line indicates 1:1 ratio."}


p <- ggplot(datxy, aes(x=as.factor(YEAR),y=ratio, color=SURVEY))
p <- p + geom_boxplot(width=1, position = position_dodge(width = 1))
p <- p + geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="year", y="iRec:creel")
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="Area", y="count")

p2 <- ggplot(datxy, aes(x=as.factor(YEAR),y=ratio, color=SURVEY))
p2 <- p2 + geom_boxplot(width=1, position = position_dodge(width = 1))
p2 <- p2 + geom_hline(aes(yintercept=1),size=1.1, alpha=.5)
p2 <- p2 + scale_color_viridis_d(end = 0.8,option = "A")
p2 <- p2 + theme_bw(16)+labs( x="year", y="iRec:creel")
p2 <- p2 + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="Area", y="count")
p2 <- p2 + coord_cartesian( ylim = c(0,15))


grid.newpage()
grid.draw(rbind(ggplotGrob(p), ggplotGrob(p2), size = "last"))


```



# Data distribution

Creel estimates are incomplete. How would there missing values affect the model fit? 

## Observation counts by month

```{r disbymonth, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = "Observation counts for creel and iRec by month. Note: iRec values were inputed, as the absence of a response implies zero catch. "}



p <- ggplot(dat)
p <- p + geom_bar(aes(x=as.factor(MONTH), color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="month", y="count")
p
```

## Observation counts by area

```{r disbyarea, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = "Observation counts for creel and iRec by area. Note: iRec values were inputed, as the absence of a response implies zero catch. "}

p <- ggplot(dat)
p <- p + geom_bar(aes(x=as.factor(AREA), color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="Area", y="count")
p

```

## Observation counts by year

```{r disbyyear, echo=FALSE, fig.height = 6, fig.width = 9, fig.cap = "Observation counts for creel and iRec by year. Note: iRec values were inputed, as the absence of a response implies zero catch. "}

p <- ggplot(dat)
p <- p + geom_bar(aes(x=as.factor(YEAR), color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="year", y="count")
p

```


## sum of estimates by month

```{r estbymonth, echo=FALSE, fig.height = 7, fig.width = 12, fig.cap = "Sum of estimates for creel and iRec by month. Note: iRec values were inputed, as the absence of a response implies zero catch. "}

datmon <- dat %>%
group_by( MONTH,DISPOSITION,SURVEY) %>% summarise(ESTIMATE = sum(ESTIMATE))


p <- ggplot(datmon)
p <- p + geom_col(aes(x=as.factor(MONTH),y=ESTIMATE, color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + facet_wrap(~DISPOSITION)
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="month", y="sum(obs)")
p
```

## sum of estimates by area

```{r estbyarea, echo=FALSE, fig.height =9, fig.width = 20, fig.cap = "Sum of estimates for creel and iRec by area. Note: iRec values were inputed, as the absence of a response implies zero catch."}

datmon <- dat %>%
group_by( AREA,DISPOSITION,SURVEY) %>% summarise(ESTIMATE = sum(ESTIMATE))


p <- ggplot(datmon)
p <- p + geom_col(aes(x=as.factor(AREA),y=ESTIMATE, color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + facet_wrap(~DISPOSITION)
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="area", y="sum(obs)")
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="Area", y="count")
p

```


## sum of estimates by year

```{r estbyyear, echo=FALSE, fig.height =9, fig.width = 20, fig.cap = "Sum of estimates for creel and iRec by year. Note: iRec values were inputed, as the absence of a response implies zero catch."}

datyr<- dat %>%
group_by( YEAR,DISPOSITION,SURVEY) %>% summarise(ESTIMATE = sum(ESTIMATE))


p <- ggplot(datyr)
p <- p + geom_col(aes(x=as.factor(YEAR),y=ESTIMATE, color=SURVEY, fill=SURVEY),alpha=.5, position=position_dodge2(width = 0.9, preserve = "single"))
p <- p + facet_wrap(~DISPOSITION)
p <- p + scale_fill_viridis_d(end = 0.8,option = "A")
p <- p + scale_color_viridis_d(end = 0.8,option = "A")
p <- p + theme_bw(16)+labs( x="year", y="sum(obs)")
p <- p + theme(axis.text.x = element_text(angle = 45, hjust=1))+labs( x="Area", y="count")
p

```


# Background

We propose to develop a calibration model to combine both survey methods to a single recreational catch estimate, taking into account month and area. 

I am thinking this would be a state-space mode as we have error associated with both surveys, something like:


The simple model (one currently being used): 

\begin{align}
&iREC_i =  \beta_1\cdot z_i + \epsilon_i\\
&z_i = Creel_i   + \nu_{i} 
\end{align}

\begin{align}
&\epsilon_i \sim N(0,\sigma_{\epsilon})\\
&\nu_{i} \sim N(0,\sigma_{\nu})
\end{align}

 where:

$iREC_i$ =  iREC observation 

$Creel_i$ =  creel observation 

$\beta_1$ =  slope associated with $Creel_i$ 

$z_i$ = smoothed ('true') $Creel_i$ 

$\epsilon_i$ and $\nu_{i}$ are random normal variables 


Adding area and month:

\begin{align}
&iREC_i =  \beta_1\cdot z_i + \beta_2\cdot area_i + \beta_3\cdot month_i + \epsilon_i\\
&z_i = Creel_i + \nu_{i} 
\end{align}

\begin{align}
&\epsilon_i \sim N(0,\sigma_{\epsilon})\\
&\nu_{i} \sim N(0,\sigma_{\nu})
\end{align}

 where: 
$\beta_2$ = slope associated with area 



$area_i$ = area for each observation (categorical) 

$\beta_3$ = slope associated with month 

$month_i$ = month for each observation (cyclic variable?) 

## Questions about model formulation:


**Q - Which parametes should be treated as "random"? Should I set $z_i$ as a random parameter?**

**Q- should I consider using a hyper prior on area? And treat it as a random effect?**

**Q -  I am not sure of how to account for missing months (if I have no creel data in winter months).**

**Q - Month as a cyclic variable?  Using cyclic splines? maybe not a good idea give the plot of ratios by months**

**Q - All estimates os iRec and creel have sds associated with them. We could weight the observations with cvs, as we do in survey indexes in stock assessment models. Suggestions?**


**Q - Interaction terms with area and month?**

It is true that interactions by area and month may exist. But I am not sure I have enough data to estimate it? maybe consider a full model with the interactions?

**Q - Consider year as a covariate?**

iRec response rates and response quality may have improved over the years as users became more familiar with the interface. 
I don't think we would have enough observations to consider year, area and month. But maybe if we drop month we could include year?
Or we could aggregate the years into blocks, e.g. pre and pos 2017.

**Q -  Which program to use? TMB, Stan, brms, rstanarm, or glmmTMB** 

I think we are inclined to use a Bayesian framework. Depending on who will be using this model, they might use a s ingle point or credible intervals, it would be good to have the option. 


glmmTMB - I am not sure we can get Bayesian estimates with it? maybe I'm wrong?
TMB(stan) - I already know. 
Stan,brms, rstanarm - I am not super knowledgeable of it but keen to learn. 

Basically I am flexible, but I might need more help depending on the program we choose.  


**Q - Spatial covariance?**

Can we account for spatial covariance if areas are discrete units (i.e. PFMA) and not continuous? How would we inform the model that Area 18 and 29 are neighbors?
See [the PFMA maps](https://www.pac.dfo-mpo.gc.ca/fm-gp/maps-cartes/areas-secteurs/index-eng.html)

**Q- spline on month estimates: this might not be relevant based on plots of the ratio between iRec and creel estimates. But interested in your thoughts.** 
